---
title: "Exploratory Data Analysis "
author: "Xing Song"
date: "10/18/2022"
output: html_document
---

Preparation: report setup, load packages, load data

```{r setup}
#set up rmd 
knitr::opts_chunk$set(message=F,warning=F,fig.width=6, fig.height=3)
pacman::p_load(tidyverse,knitr,kableExtra)
aset<-readRDS("C:/Users/Administrator/Documents/classHMI7001/data/final_set.rda")
```


**************************************************************************************************

## Exploratory Data Analysis (EDA)

The primary aim of EDA is to:

* **Maximize insight** into the database/understand the database structure
* **Visualize potential relationships** (direction and magnitude) between exposure and outcome variables
* **Detect outliers and anomalies** (values that are significantly different from the other observations)
* **Extract and create clinically** relevant variables.

**************************************************************************************************

### Categorical Data/Variables 

#### Univariate EDA

##### Non-Graphical Methods

*Tabulation, or Frequency table*
    
```{r}
##==============================================
# use "table","prop.table" function 
# for quick check
##==============================================
table(aset$AGE_AT_ALS1DX,dnn=list("freq"))
prop.table(table(aset$AGE_AT_ALS1DX,dnn=list("rel_freq")))

##==============================================
# use "kable" function for nicer print-out
##==============================================
uni_freq_tbl<- aset %>%
  group_by(AGE_AT_ALS1DX) %>%
  summarise(freq=n()) %>% 
  ungroup %>%
  mutate(rel_freq=round(freq/1373,4))

kable(stack_tbl,
      col.names = c("AGE_AT_ALS1DX","Frequency","Relative Frequency(%)"),
      caption="Frequency Table of different ")
```


#### Graphical Methods

*Barplot*

```{r race_barplot}
# example for nominal data
ggplot(data=idd_bmi,aes(x=DIAGNOSIS)) +
  geom_bar()

# example for ordinal data
ggplot(data=idd_bmi,aes(x=bmi_classify)) +
  geom_bar()
```


*Pie chart*
  
```{r race_pie}
ggplot(idd_bmi %>% group_by(DIAGNOSIS) %>%
         dplyr::summarize(freq=n()) %>% ungroup, 
       aes(x="",y=freq,fill=DIAGNOSIS)) +
  geom_bar(width=1,stat="identity")+
  coord_polar("y", start=0)+
  labs(x="",y="")
```

*************************************************************************************

#### Multivariate EDA

##### Non-Graphical Method

*Cross tabulation, or Contingency table, or Frequency table*
    
```{r tbl_dx_bmi}
bivar_freq_tbl<-table(idd_bmi$bmi_classify,idd_bmi$DIAGNOSIS,
      dnn=list("BMI","DIASNOSIS"))

bivar_freq_tbl
```

##### Graphical Method

*Conditional Barplot*

```{r cond_bar}
ggplot(data=idd_bmi,aes(x=DIAGNOSIS,fill=bmi_classify)) +
  geom_bar(position="dodge")
```

*Heatmap*

```{r heat_race_dx}
ggplot(idd_bmi %>% group_by(DIAGNOSIS,bmi_classify) %>%
         dplyr::summarize(freq=n()) %>% ungroup, 
       aes(x=bmi_classify,y=DIAGNOSIS,fill=freq))+
  geom_tile(colour = "white") + 
  scale_fill_gradient(low = "lightblue",high = "steelblue") + 
  theme_classic()

```

**************************************************************************************************

### Numerical Data/Variables

#### Univariate EDA

##### Non-Graphical Method

*Descriptive Statistics*
    
a) Central Tendency: mean, median, mode
        
```{r bmi_center}
#mean
mean(idd_bmi$bmi_median)
#median
median(idd_bmi$bmi_median)
#mode
idd_bmi %>% 
  group_by(RACE) %>% 
  summarise(n=n()) %>% 
  slice(which.max(n))
```

b) Spread: min, max, variance, standard deviation (sd), interquartile range (IQR)
        
```{r bmi_sprerad}
#range
max(idd_bmi$bmi_median)-min(idd_bmi$bmi_median)

#variance
var(idd_bmi$bmi_median)

#sd
sd(idd_bmi$bmi_median)

#IQR
IQR(idd_bmi$bmi_median)

#5-number summary
summary(idd_bmi$bmi_median)
```

c) Shape: skewness (normal distribution is near 0), kurtosis (normal distribution is around 3)


##### Graphical Method

*Histogram, Density plot*
    
```{r bmi_hist}
#histogram
ggplot(idd_bmi,aes(bmi_median))+
  geom_histogram(bins = 30)

#density plot
ggplot(idd_bmi,aes(bmi_median))+
  geom_density(bins = 30)
```

*Boxplot*
```{r boxplot}
#vertical
ggplot(idd_bmi,aes(y=bmi_median))+
  geom_boxplot(bins = 30)

#horizontal
ggplot(idd_bmi,aes(x=bmi_median))+
  geom_boxplot(bins = 30)
```


#### Multivariate EDA

##### Non-Graphical Method

*Pearson Correlation*
        
* The closer it is to 1, the more linearly related
* positive sign suggests positive correlation
          
```{r cor_bmi_age}
cor(idd_bmi$bmi_median,idd_bmi$AGE_AT_FIRSTDX,method="pearson")
```

*Spearman Correlation*
```{r cor_bmi_age}
cor(idd_bmi$bmi_median,idd_bmi$AGE_AT_FIRSTDX,method="spearman")
```


##### Graphical Method

*Scatter Plot*

```{r age_bmi}
ggplot(idd_bmi %>% filter(!is.na(AGE_AT_FIRSTDX)&bmi_median<50&AGE_AT_FIRSTDX>=18),
       aes(x=AGE_AT_FIRSTDX,y=bmi_median))+
  geom_point()+
  geom_smooth(method="loess")
```


#### Mixture Data

##### Non-Graphical Method

*ANOVA*

```{r anova_bmi_dx}
fit_aov<-aov(bmi_median ~ DIAGNOSIS, data=idd_bmi)
summary(fit_aov)
```

##### Graphical Method

*Conditional Boxplot*

```{r bmi_boxplot}
#vertical boxplot
ggplot(idd_bmi,aes(y=bmi_median,fill=DIAGNOSIS))+
  geom_boxplot()
```


*************************************************************

