---
title: "Exploratory Data Analysis "
author: "Xing Song"
date: "10/18/2022"
output: html_document
---

Preparation: report setup, load packages, load data

```{r setup}
#set up rmd 
knitr::opts_chunk$set(message=F,warning=F,fig.width=6, fig.height=3)
pacman::p_load(tidyverse,
               kableExtra)
aset<-readRDS("C:/Users/Administrator/Documents/classHMI7001/data/final_set.rda")
```


**************************************************************************************************

## Exploratory Data Analysis (EDA)

EDA:  refers to a set of procedures for producing descriptive and graphical summary of the data. The term is developed by John Tukey: “The best single device for suggesting, and at times answering questions beyond those originally posed is the graphical display”. The primary aim of EDA is to:

* **Maximize insight** into the database/understand the database structure
* **Visualize potential relationships** (direction and magnitude) between exposure and outcome variables
* **Detect outliers and anomalies** (values that are significantly different from the other observations)
* **Extract and create clinically** relevant variables.

**************************************************************************************************

### Categorical Data/Variables 
```{r}
aset<-aset %>%
  mutate(AGEGRP=case_when(AGE_AT_ALS1DX<65~'AgeGrp1',
                          AGE_AT_ALS1DX>=65&AGE_AT_ALS1DX<70~'AgeGrp2',
                          AGE_AT_ALS1DX>=70&AGE_AT_ALS1DX<75~'AgeGrp3',
                          AGE_AT_ALS1DX>=75&AGE_AT_ALS1DX<80~'AgeGrp4',
                          TRUE ~'AgeGrp5'))

head(aset)
```
#### Univariate EDA

##### Non-Graphical Methods

*Tabulation, or Frequency table*
    
```{r}
##==============================================
# use "table","prop.table" function 
# for quick check
##==============================================
table(aset$RACE,dnn=list("freq"))
prop.table(table(aset$RACE,dnn=list("rel_freq")))
```


```{r}
##==============================================
# use "kable" function for nicer print-out
##==============================================
uni_freq_tbl<- aset %>%
  group_by(RACE) %>%
  summarise(freq=n()) %>% 
  ungroup %>%
  mutate(rel_freq=round(freq/1373,4))

kable(uni_freq_tbl,
      col.names = c("RACE","Frequency","Relative Frequency(%)"),
      caption="Frequency Table of Racial Groups")
```


#### Graphical Methods

*Barplot*

```{r}
# example for nominal data
ggplot(data=aset,aes(x=RACE)) +
  geom_bar()
```


```{r}
# example for ordinal data
ggplot(data=aset,aes(x=AGEGRP)) +
  geom_bar()
```


*Pie chart*
  
```{r}
ggplot(aset %>% group_by(RACE) %>%
         dplyr::summarize(freq=n()) %>% ungroup, 
       aes(x="",y=freq,fill=RACE)) +
  geom_bar(width=1,stat="identity")+
  coord_polar("y", start=0)+
  labs(x="",y="")
```


*************************************************************************************

#### Multivariate EDA

##### Non-Graphical Method

*Cross tabulation, or Contingency table, or Frequency table*
    
```{r}
bivar_freq_tbl<-table(aset$RACE,aset$AGEGRP,
      dnn=list("RACE","AGEGRP"))

bivar_freq_tbl
```

```{r}
prop.table(bivar_freq_tbl)
```

```{r}
# row proportions
prop.table(bivar_freq_tbl,margin=1)
```

```{r}
# column proportions
prop.table(bivar_freq_tbl,margin=2)
```


##### Graphical Method

*Conditional Barplot*

```{r}
ggplot(data=aset,aes(x=RACE,fill=AGEGRP)) +
  geom_bar(position="dodge")
```

*Heatmap*

```{r heat_race_dx}
ggplot(aset %>% group_by(RACE,AGEGRP) %>%
         dplyr::summarize(freq=n()) %>% ungroup, 
       aes(x=AGEGRP,y=RACE,fill=freq))+
  geom_tile(colour = "white") + 
  scale_fill_gradient(low = "lightblue",high = "steelblue") + 
  theme_classic()

```

**************************************************************************************************

### Numerical Data/Variables

#### Univariate EDA

##### Non-Graphical Method

*Descriptive Statistics*
    
a) Central Tendency: mean, median, mode
        
```{r}
#mean
mean(aset$AGE_AT_ALS1DX)
#median
median(aset$AGE_AT_ALS1DX)
#mode
aset %>% 
  group_by(RACE) %>% 
  summarise(n=n()) %>% 
  slice(which.max(n))
```

b) Spread: min, max, variance, standard deviation (sd), interquartile range (IQR)
        
```{r}
#range
max(aset$AGE_AT_ALS1DX)-min(aset$AGE_AT_ALS1DX)

#variance
var(aset$AGE_AT_ALS1DX)

#sd
sd(aset$AGE_AT_ALS1DX)

#IQR
IQR(aset$AGE_AT_ALS1DX)

#5-number summary
summary(aset$AGE_AT_ALS1DX)
```

c) Shape: skewness (normal distribution is near 0), kurtosis (normal distribution is around 3)
```{r}
pacman::p_load("moments")

# skewness
skewness(aset$AGE_AT_ALS1DX)

# kurtosis
kurtosis(aset$AGE_AT_ALS1DX)

```

##### Graphical Method

*Histogram, Density plot*
    
```{r}
#histogram
ggplot(aset,aes(AGE_AT_ALS1DX))+
  geom_histogram(bins = 30)

#density plot
ggplot(aset,aes(AGE_AT_ALS1DX))+
  geom_density(bins = 30)
```

*Boxplot*
```{r}
#vertical
ggplot(aset,aes(y=AGE_AT_ALS1DX))+
  geom_boxplot(bins = 30)

#horizontal
ggplot(aset,aes(x=AGE_AT_ALS1DX))+
  geom_boxplot(bins = 30)
```


#### Multivariate EDA

##### Non-Graphical Method

*Pearson Correlation*
        
* The closer it is to 1, the more linearly related
* positive sign suggests positive correlation
          
```{r}
cor(aset$AGE_AT_ALS1DX,aset$RILUZ_DUR,method="pearson")
```

*Spearman Correlation*
```{r}
cor(aset$AGE_AT_ALS1DX,aset$RILUZ_DUR,method="spearman")
```


##### Graphical Method

*Scatter Plot*

```{r}
ggplot(aset, aes(x=AGE_AT_ALS1DX,y=RILUZ_DUR))+
  geom_point()+
  geom_smooth(method="loess",formula='y ~ x')
```


#### Mixture Data

##### Non-Graphical Method

*ANOVA*

```{r}
fit_aov<-aov(AGE_AT_ALS1DX ~ SEX, data=aset)
summary(fit_aov)
```

##### Graphical Method

*Conditional Boxplot*

```{r bmi_boxplot}
#vertical boxplot
ggplot(aset,aes(y=AGE_AT_ALS1DX,fill=SEX))+
  geom_boxplot()
```
