---
title: "Data Preparation and Preprocessing"
author: "Xing Song"
date: "10/06/2022"
output: html_document
---

Let's do some preparation for the class by running the following chunk of codes:

```{r setup, include=FALSE}
#set up rmd print-out style
knitr::opts_chunk$set(message=F,
                      warning=F,
                      fig.width=6, 
                      fig.height=3)

#install all the packages for this script
pacman::p_load(tidyverse,# data cleaning packages
               finalfit,mice # missing data detection and imputation
               )
```

```{r}
pt_tbl<-readRDS("C:/Users/Administrator/Documents/classHMI7001/data/als_pt_table.rda")
riluz_tbl<-readRDS("C:/Users/Administrator/Documents/classHMI7001/data/als_riluzole.rda")
```

## Quantitative and Qualitative Data (11.2.3)

#### Qualitative Data (categorical data)

**nominal data**

```{r nom_data}
ggplot(pt_tbl,aes(x=RACE))+
  geom_histogram(stat="count")
```

**ordinal**

```{r ord_data}
##==============================================
# "mutate" function for adding/altering columns
# "case_when" function
##==============================================
#add a new column, "age_group", that groups ages into 5 groups
pt_tbl_add_agegrp<-pt_tbl %>%
  mutate(age_group=case_when(AGE_AT_ALS1DX<65~'Group1',
                             AGE_AT_ALS1DX>=65&AGE_AT_ALS1DX<70~'Group2',
                             AGE_AT_ALS1DX>=70&AGE_AT_ALS1DX<75~'Group3',
                             AGE_AT_ALS1DX>=75&AGE_AT_ALS1DX<80~'Group4',
                             AGE_AT_ALS1DX>=80~'Group5'))
#print first 6 rows
head(pt_tbl_add_agegrp)

#plot out the age group
ggplot(pt_tbl_add_agegrp,aes(x=age_group))+
  geom_bar(stat="count")
```



**********************************                     

#### Quantitative Data (numerical data)

**discrete data**

```{r cont_dat}
ggplot(pt_tbl,aes(x=AGE_AT_ALS1DX))+
  geom_histogram(fill="blue",bins = 40)

#5-number summary
table(pt_tbl$AGE_AT_ALS1DX)
```

**continuous data**

```{r cont_dat}
ggplot(pt_tbl,aes(x=AGE_AT_ALS1DX))+
  geom_density(fill="blue")

#5-number summary
summary(pt_tbl$AGE_AT_ALS1DX)
```

**************************************************************************************************
**************************************************************************************************

## Data Preprocessing

General steps for data preprocessing are **Data Integration**, **Data Abstraction**, **Data Cleaning**, **Data Transformation** and **Data Reduction**. 

### Data Abstraction
As you can see, in the `riluzole_tbl`, there are multiple medication entries for the drug riluzole. To get an analytic set with **one patient observation per row**, we need to perform the **data abstraction** step.

```{r}
##======================================================
# "group_by" and "summarize" function for summarizing 
#  variables according to certain grouping criteria and
#  functions of interests
##======================================================
riluz_tbl_dur<-riluz_tbl %>%
  group_by(PATID) %>%
  summarize(riluz_begin = min(RX_ORDER_DATE),
            riluz_end = max(RX_START_DATE),
            .groups = "drop")
```


```{r}
##=========================================================================================
# "ymd" and "difftime" functions from "lubridate" package to calculate date difference
# The "lubridate" package in tidyverse provides functions for operations on dates and times
# more on: https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html
##=======================================================================================
riluz_tbl_dur<-riluz_tbl_dur %>%
  mutate(RILUZ_DUR = difftime(lubridate::ymd(riluz_end), lubridate::ymd(riluz_begin), units = "days")) %>%
  mutate(RILUZ_DUR = as.numeric(RILUZ_DUR) + 1)
```


**************************************************************************************************
**************************************************************************************************

### Data Integration

A good strategy for data integration is to conceptualize what you want your final analytic dataset to look like even before data collection, guided by the following questions:
* will the final analytic dataset be **one patient per row** or **one encounter per row** or **one particular event per row**?
* what will be the layout of the final analytic dataset? More specifically, what are the columns in the final analytic dataset? 

Let's form a simple analytic dataset with: 
1. one patient per row;
2. layout to look like: 

|PATID|SEX|RACE|AGE_AT_ALS1DX|RILUZ_DUR|RILUZ_IND|DAYS_ALS1DX_TO_DEATH|DAYS_ALS1DX_TO_CENSOR|MORT_IND |MORT_1YR |
|1234 |M  |05  |70           |300      |1        |NA                  |200                |0        |0        |

```{r}
##========================================================
# "left_join" or "inner_joint function to integrate
# multiple tables together
##========================================================
final_set<-pt_tbl %>%
  left_join(riluz_tbl_dur,by="PATID") %>%
  mutate(RILUZ_IND = case_when(!is.na(RILUZ_DUR) ~ 1,
                              TRUE ~ 0),
         MORT_IND = case_when(!is.na(DAYS_ALS1DX_TO_DEATH) ~ 1,
                              TRUE ~ 0),
         MORT_1YR = case_when(DAYS_ALS1DX_TO_DEATH<=365 ~ 1,
                              is.na(DAYS_ALS1DX_TO_DEATH) & DAYS_ALS1DX_TO_CENSOR<=365 ~ NA_real_,
                              TRUE ~ 0)) %>%
  select(PATID,SEX,RACE,AGE_AT_ALS1DX,RILUZ_DUR,RILUZ_IND,DAYS_ALS1DX_TO_DEATH,DAYS_ALS1DX_TO_CENSOR,MORT_IND, MORT_1YR)
```


**************************************************************************************************
**************************************************************************************************

### Data Cleaning 

#### Data Cleaning - Missing Data 

```{r}
str(final_set)
```

```{r}
##========================================================
# "finalfit" package provides functions that help you quickly create 
# elegant final results tables and plots, including an effective
# way to look at missing patterns
# 
# more on: https://finalfit.org/
##======================================================== 
explanatory = c("RACE","RILUZ_DUR")
dependent = 'MORT_IND'
final_set %>%
  missing_pattern(dependent,explanatory)
```



```{r}
##=====================================================
# "replace_na" function for quick imputation using
# a constant value
##=====================================================
final_set<-final_set %>%
  replace_na(list(RACE='UN'))

final_set %>%
  missing_pattern(dependent,explanatory)
```


### Data Transformation

#### One-hot encoding 
```{r}
race_ohe<-final_set %>% 
  dplyr::select(PATID,RACE) %>%
  mutate(ind=1) %>%
  spread(RACE,ind, fill = 0)

head(race_ohe)
```
