---
title: "Survival Analysis"
author: "Xing Song"
date: '2022-11-10'
output: html_document
---

Let's do some preparation for the class by running the following chunk of codes

```{r setup}
#set up rmd 
knitr::opts_chunk$set(message=F,
                      warning=F,
                      echo=F,
                      fig.width=6,
                      fig.height=4)

#load packages
pacman::p_load(tidyverse,
               survival,
               survminer,
               gtsummary)

#load data
aset<-readRDS("C:/Users/Administrator/Documents/classHMI7001/data/final_set.rda") %>%
  mutate(SEX_F = case_when(SEX=='F' ~ 1,
                           TRUE ~ 0)) %>%
  mutate(status=MORT_IND,
         time=coalesce(DAYS_ALS1DX_TO_DEATH,DAYS_ALS1DX_TO_CENSOR))
```


## Kaplan-Meier Curve

Let's first get familiar with survival object. The `Surv()` function creates a survival object that contains three parts: a numerical part that is the survival time; a `+` or `:1` symbol suggesting the subject being censored or an event. For example,

```{r}
Surv(aset$time,aset$status)[1:10]
```

```{r}
km_fit<-survfit(Surv(time,status) ~ 1, data = aset)
```

### Estimate ð‘¥-year survival, or life table

```{r}
summary(km_fit,times = 365.25*c(1:5))
```


```{r}
ggsurvplot(fit = km_fit,
           pval = TRUE, conf.int = TRUE,risk.table = TRUE,
           surv.median.line = "hv",linetype = "strata", 
           xlab = "Days", break.x.by = 365*2,
           ylab = "Mortality Endpoint")
```

### Estimate median survival time

```{r}
survfit(Surv(time, status) ~ 1, data = aset) %>% 
  tbl_survfit(
    probs = 0.5,
    label_header = "**Median survival (95% CI)**"
  )
```


### Stratified KM curves

```{r,fig.height=8,fig.width=12}
ggsurvplot(
    fit = survfit(Surv(time,status) ~ RILUZ_IND, data = aset),
    pval = TRUE, conf.int = TRUE,risk.table = TRUE,
    surv.median.line = "hv",linetype = "strata", 
    xlab = "Days", break.x.by = 365,
    ylab = "Mortality Endpoint")
```


```{r,fig.height=8,fig.width=12}
ggsurvplot(
    fit = survfit(Surv(time,status) ~ SEX_F, data = aset),
    pval = TRUE, conf.int = TRUE,risk.table = TRUE,
    surv.median.line = "hv",linetype = "strata", 
    xlab = "Days", break.x.by = 365,
    ylab = "Mortality Endpoint")
```


## Cox Proportional Hazard Regression Model

```{r}
ggplot(aset,aes(x=time))+
  geom_density(aes(fill=as.factor(status)),alpha=0.5)
```


```{r}
cox_fit<-coxph(Surv(time, status) ~ RILUZ_IND, data = aset)
summary(cox_fit)
```


```{r}
coxph(Surv(time, status) ~ RILUZ_IND, data = aset) %>% 
  tbl_regression(exp = TRUE)
```


```{r}
coxph(Surv(time, status) ~ SEX_F, data = aset) %>% 
  tbl_regression(exp = TRUE)
```

```{r}
cox_fit2<-coxph(Surv(time, status) ~ AGE_AT_ALS1DX + RILUZ_IND + RILUZ_DUR , data = aset)
summary(cox_fit2)
```


### Model Diagnostics

```{r}
res<-cox.zph(cox_fit2)
plot(res)
```


## Accelerated Failure Time Model

```{r}
aft_fit1<-survreg(Surv(time, status) ~ AGE_AT_ALS1DX + RILUZ_IND + RILUZ_DUR, data = aset,
              dist = "exp")
summary(aft_fit1)
```


```{r}
aft_fit2<-survreg(Surv(time, status) ~ AGE_AT_ALS1DX + RILUZ_IND + RILUZ_DUR, data = aset,
              dist = "weib")
summary(aft_fit2)
```


```{r}
aft_fit3<-survreg(Surv(time, status) ~ AGE_AT_ALS1DX + RILUZ_IND + RILUZ_DUR, data = aset,
              dist = "logl")
summary(aft_fit3)
```


```{r}
aft_fit4<-survreg(Surv(time, status) ~ AGE_AT_ALS1DX + RILUZ_IND + RILUZ_DUR, data = aset,
              dist = "logn")
summary(aft_fit4)
```


```{r}
AIC(aft_fit1,
    aft_fit2,
    aft_fit3,
    aft_fit4,
    cox_fit2)
```
